---
title: "Epigenetic Age in Eye Tissue"
author: "Kendrix"
date: "12/1/2020"
output:
  html_document:
      highlight: pygments
      theme: united
      toc: true
      toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Aim: To test the feasibility of testing the epigenetic clocks on the eye tissue using all publicly available control eye datasets.

Pre-processing script adapted from Chaini Konwar.


---------------------------------------------------------------------------------------------------------------------------

```{r, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
library(GEOquery)
library(RCurl)
library(GEOmetadb)
library(dendextend)
library(ArrayExpress)
library(methylumi)
library(lumi)
library(lattice)
library(gplots)
library(RColorBrewer)
library(limma)
library(ROC)
library(matrixStats)
library(reshape)
library(sva)
library(grid)
library(gridExtra)
library(ape)
library(Hmisc)
library(RCurl)
library(wateRmelon)
library(minfiData)
library(minfi)
library(robustHD)
library(ewastools)
library(omicsPrint)
library(doParallel)
library(jcolors)
library(plyr)
library(tidyverse)
library(ggrepel)
library(ggpubr)
library(IlluminaHumanMethylationEPICanno.ilm10b2.hg19)
library(IlluminaHumanMethylationEPICmanifest)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(IlluminaHumanMethylation450kmanifest)
library(FlowSorted.Blood.EPIC)
library(FlowSorted.Blood.450k)
library(FlowSorted.CordBloodCombined.450k)
library(Biobase)
library(data.table)
library(factoextra)
library(Metrics)
library(quantro)
library(impute)

setwd("~/KoborLab/kobor_space/kendrix/macular_degeneration/")
```

---------------------------------------------------------------------------------------------------------------------------


##1.0 Data exploration and object creation


###1.1 Download eye datasets from public databases.

Using all publicly eye dataset, combine all eye samples into a single large dataset.

As of Dec 2020, 

- There are 6 methylation by microarray datasets in GEO, 4 are cell lines and 2 are usable datasets that are also available in ArrayExpress. They are listed below. 

- There are 2 methylation by microarray datasets in ArrayExpress, which are similar to the ones in GEO. 


Datasets used:
**GEO**: GSE102952, GSE57362
**ArrayExpress**: E-MTAB-7183

Exclusion criteria: Samples with retinal diseases.


```{r}
#GEO dataset: GSE102952

#Download supplementary data.
getGEOSuppFiles("GSE102952", fetch_files = TRUE, baseDir = "~/KoborLab/kobor_space/kendrix/macular_degeneration/data/")
untar("~/KoborLab/kobor_space/kendrix/macular_degeneration/data/GSE102952/GSE102952_RAW.tar", exdir = "/home/BCRICWH.LAN/kendrix.kek/KoborLab/kobor_space/kendrix/macular_degeneration/data/GSE102952/")
file.remove("~/KoborLab/kobor_space/kendrix/macular_degeneration/data/GSE102952/GSE102952_RAW.tar")

#Download GSE102952 dataset.
Oliver_AMD <- getGEO("GSE102952", GSEMatrix = TRUE, destdir = "~/KoborLab/kobor_space/kendrix/macular_degeneration/data/GSE102952")
show(Oliver_AMD)

#Unzip all files in the folder.
head(list.files("~/KoborLab/kobor_space/kendrix/macular_degeneration/data/GSE102952", pattern = ".gz$"))
suppFiles <- list.files("~/KoborLab/kobor_space/kendrix/macular_degeneration/data/GSE102952", pattern = "gz$", full = TRUE)
sapply(suppFiles, gunzip, overwrite = TRUE)

#Get pData table.
Oliver_AMD.pData <- pData(phenoData(Oliver_AMD[[1]]))


#Get beta matrix.
Oliver_AMD.betas <- read.table("~/KoborLab/kobor_space/kendrix/macular_degeneration/data/GSE102952/GSE102952_series_matrix.txt", header = TRUE, fill = TRUE, skip = 63, stringsAsFactors = FALSE)
rownames(Oliver_AMD.betas) <- Oliver_AMD.betas$ID_REF
Oliver_AMD.betas <- Oliver_AMD.betas[,-1]

#Make sample names consistent. 
identical(rownames(Oliver_AMD.pData), colnames(Oliver_AMD.betas)) #TRUE.
colnames(Oliver_AMD.pData)[1] <- "Sample_Name"

#Fix sample names of pData and beta matrix.
rownames(Oliver_AMD.pData) <- Oliver_AMD.pData$Sample_Name
colnames(Oliver_AMD.betas) <- Oliver_AMD.pData$Sample_Name
identical(rownames(Oliver_AMD.pData), colnames(Oliver_AMD.betas)) #TRUE. 

#Check for NAs and infinite values.
sum(is.na(Oliver_AMD.betas)) #1.
```




